Argennon: A Scalable Smart Contract Platform

### 2022-04-02 tex/sections/avm.tex
Fix some readability issues
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and requirements of the Argennon platform.
Argennon supports two standard representations: one is a high level text based language which needs
costly compilation before being executed on a hardware machine. The other is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving the high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. On the other hand, the low level language is designed for
efficient execution of applications that are not frequently used.

@@

The AscEE executes the requests contained in each block of the Argennon
blockchain in a three-step procedure. The first step is the \emph{preprocessing step}. In this step, the required
data structures for executing requests are constructed. This step is
designed in a way that can be done fully in parallel for each request without any risks of data races. The second
step is the \emph{Data Dependency Analysis (DDA) step}.
In this step by analyzing data dependency between requests, the AscEE determines requests that can be run in parallel
and requests that need to be run sequentially. This information is represented using an \emph{execution dag} data
structure. In the final step, request are executed using this data structure and helper data structures built in the
first step.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing HTTP requests, producing
their HTTP response and updating related heap chunks. The AscEE executes requests sequentially \footnote{Actually
requests are executed in parallel but by performing data dependency analysis the result is guaranteed to be identical
with sequential execution of requests.} and each request has a separate \emph{execution session}. Execution sessions
are separate sessions of executing smart contract's code in order to fulfill \emph{external} HTTP requests. An
external request is a request that is not made by another Argennon application.

The state and data of an execution session will be
destroyed at the end of the session and only the state of heap chunks is preserved. If a session fails and does not
complete normally, it will not have any effect on any heap chunks.

During an execution session an application can make \emph{internal} HTTP requests to other applications. Those
requests will not start a new execution session and will be executed within the current session. In AscEE making a
HTTP request to an application is similar to a function invocation. For that reason, we also refer to internal requests
as application calls.

The AscEE is designed based on \emph{optional decoupling principle}. When an application makes a request to another
application, optionally it can request to be decoupled from the called application. That would mean the called
application could not affect its caller's state by reentrancy, or could not fail the session by using excessive
resources or performing illegal operations.

@@

In Argennon a unique identifier is assigned to every application, heap chunk and account. Therefore, three distinct
identifier types exist: \texttt{appID}, \texttt{accountID}, and \texttt{chunkID}.
All these identifiers are \emph{prefix codes}, and hence can be represented by
\emph{prefix trees}\footnote{Also called tries.}.

@@

\begin{itemize}
    \item \texttt{appID} is a prefix code built by \emph{applications} prefix tree. An application ID cannot
    be \texttt{0x0}.

    \item \texttt{accountID} is a prefix code built by \emph{accounts} prefix tree. An account ID cannot
    be \texttt{0x0} or \texttt{0x1}.

@@

The persistent data of an Argennon application is stored in heap chunks. An application may have several heap chunks
with different sizes, and can remove or
resize its chunks or allocate new chunks. Every chunk belongs to exactly one application. Only the owner application can
modify a chunk but there is no restrictions for reading a chunk.

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself \footnote{provided it has a correct format}. This
is an important feature of the AscEE's heap, which allows applications to use the AscEE's heap as a dictionary (map)
data structure.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily find unique identifiers for its chunks.

A heap chunk can be considered as a continuous
array of bytes. Every chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated during the session like a normal memory location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the blockchain and is proposed by the block
proposer. \texttt{sizeUpperBound} indicates the upper bound of the value of \texttt{chunkSize} for a chunk in a block.
A validator can use this value to safely allocate memory for a chunk.

The value of \texttt{chunkSize} at the end of the execution session will determine if a memory location at an
offset is persistent or not: Offsets lower than the chunk size are persistent, and higher offsets are not.
Non-persistent locations will be re-initialized with zero, at the start of every execution session.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is defined on a chunk
and has an offset and a size and determines accessible memory locations inside a chunk. Multiple access blocks can
be defined on a single chunk, but they must be non overlapping. Access blocks are byte addressable and can have
different
access types:

\begin{itemize}
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{additive}: only allows addition operations without overflow checking. Note that the content of these
    access blocks cannot be read.
\end{itemize}

Chunks are intended for simplifying proof checking of the data stored in the Argennon cloud and access blocks are
required for better parallelization of the request execution. An application should put the data it predicts is
needed for validating a block in the same chunk and the data it predicts is needed in a single execution
session in the same access block.

@@

An application can load a chunk with a valid prefix identifier even if that chunk does not exist. For a non-existent
chunk the value of \texttt{chunkSize} is always zero.

@@

Attachment of a request is a list of request identifiers of the current block that are "attached" to the request.
That means, for validating that request a validator first needs to "inject" the digest of attached requests into the
HTTP request.

The main usage of this feature is for fee payment. A request that wants to pay the fees for a number of requests
declares those requests as its attachments. For paying fees the payer signs the digest of requests for which he
wants to pay fees. After injecting the digest of those request by validators, that signature can be validated
correctly and securely.

@@

The AscEE uses \emph{authenticated message passing} for authorizing operations. In this method every execution
session has a set of authenticated messages. Applications explicitly pass those messages for authorizing operations.
Authentication of messages can be done using any type of cryptographic signature scheme in the preprocessing phase of
request execution. For example, we could use BLS aggregate signature to authenticate all messages of a block in bulk.

Moreover, application use built-in functions of AscEE to generate authenticated messages in run-time. This enables an
application to authorize an operation even for an application that is not directly called.

\note{Authorization by explicit messages and signatures eliminates the need for approval methods or call back patterns.}

@@

\note{The Argennon Execution Environment has no instructions for issuing cryptographic signatures.}

@@

The Argennon Smart Contract Execution Environment provides optional low level reentrancy protection by providing low
level \emph{entrance locks}. When an application acquires an entrance lock it cannot acquire that lock again and trying
to do so will result in a revert. The entrance lock of an application will be released when the application explicitly
releases its entrance lock or when the call that had acquired the lock completes.

@@

Obviously resource monitoring is easier and more efficient for optimistic sessions. However, if
an optimistic session violates its resource caps, determining the point of failure requires precise resource
measurement. Note that for implementing optional decoupling principle, we need to determine the exact application
which has failed in a call chain. For example assume that an optimistic session containing an application call
violates a 2 milliseconds execution time cap. If in the caller application the call happens exactly after 2
milliseconds of execution, a small fluctuation in the execution time measurement can
change the point of failure between the called and the main application. This can introduce nondeterministic
behaviour which could make block validation impossible. On the other hand, for a monitored session this is not an issue
because resource measurements and caps are defined per application, and we can make sure that resource caps are not too
small and can be measured easily.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    the amount of cpu time that is required for executing a session or an application call. The execution time of a
    session is measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of cpu time needed for executing a predefined standard application which is used for benchmarking a host's
    performance.

@@

    Monitored sessions perform per application call cpu-time measurement, and every application call during a monitored
    session has a separate \texttt{maxClocks} value. This value determines the maximum amount of time that the cpu
    can be used for executing that particular application call. It should be noted that the cpu timer is paused when
    the application makes a call to another application, and is resumed when the control returns. An application call
    needs at least 100 clocks and if the value of its \texttt{maxClock} is lower than this value the call will be
    considered a failed call.

@@

    Optimistic sessions measure local memory usage per session and enforce a protocol-defined cap on the total amount of
    used local memory in the session. Monitored sessions measure local memory usage per
    application call and enforce a protocol defined-cap per application call. An application call which tries to use
    more local memory than the cap will fail.
    \item \textbf{heap access list}:
    every session can only access heap locations that are declared in its access list. In addition,
    resizing heap chunks can only be done in the range of the pre-declared lower bound and upper bound.
    \item \textbf{applications list}:
    a session may only make requests to applications that are declared in its application list.
    \item \textbf{call depth}:
    during a session the number of nested application calls can not be more than a threshold. This threshold is
    determined by the Argennon protocol.
    \item \textbf{active differed calls}:
    \item \textbf{virtual signatures}:
    \item \textbf{number of entrance locks}:
\end{itemize}

The Argennon protocol allows AscEE's implementations to have at max \%50 error in measuring execution time and at
max \%25 error in measuring local memory usage. Other resources must be measured precisely.

>>>>

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and requirements of the Argennon platform.
Argennon supports two standard representations: one is a high level text based language which needs
costly compilation before being executed on a hardware machine. The other is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving the high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. On the other hand, the low level language is designed for
efficient direct execution of applications that are not frequently used.

@@

The AscEE executes the requests contained in each block of the Argennon
blockchain in a three-step procedure. The first step is the \emph{preprocessing step}. In this step, the required
data structures for executing requests are constructed. This step is
designed in a way that can be done fully in parallel for each request without any risks of data races. The second
step is the \emph{Data Dependency Analysis (DDA) step}.
In this step by analyzing data dependency between requests, the AscEE determines requests that can be run in parallel
and requests that need to be run sequentially. This information is represented using an \emph{execution dag} and in
the final step, request are executed using this data structure.

@@

The Argennon Smart Contract Execution Environment can be seen as a machine for executing Argennon applications to
fulfill HTTP requests, produce their HTTP responses and update related heap chunks. The AscEE executes requests
sequentially \footnote{Actually requests are executed in parallel but by performing data dependency analysis the
result is guaranteed to be identical
with sequential execution of requests.} and each request has a separate \emph{execution session}. Execution sessions
are separate sessions of executing smart contract's code in order to fulfill \emph{external} HTTP requests. An
external request is a request that is not made by another Argennon application.

The state of an execution session will be
destroyed at the end of the session and only the state of heap chunks is preserved. If a session fails and does not
complete normally, it will not have any effect on any heap chunks.

During an execution session an application can make \emph{internal} HTTP requests to other applications. Those
requests will not start a new execution session and will be executed within the current session. In AscEE making an
HTTP request to an application is similar to a function invocation. For that reason, we also refer to internal requests
as application calls.

The AscEE is designed based on \emph{optional decoupling principle}. When an application makes a request to another
application, optionally it can choose to be decoupled from the called application. That would mean the called
application could not affect its caller's state by reentrancy, or could not fail the session by using excessive
resources or performing illegal operations.

@@

In Argennon a unique identifier is assigned to every application, heap chunk and account. Consequently, three distinct
identifier types exist: \texttt{appID}, \texttt{accountID}, and \texttt{chunkID}.
All these identifiers are \emph{prefix codes}, and hence can be represented by
\emph{prefix trees}\footnote{Also called tries.}.

@@

\begin{itemize}
    \item \texttt{appID} is a prefix code built by \emph{applications} prefix tree. An \texttt{appID} cannot
    be \texttt{0x0}.

    \item \texttt{accountID} is a prefix code built by \emph{accounts} prefix tree. An \texttt{accountID} cannot
    be \texttt{0x0} or \texttt{0x1}.

@@

The persistent data of an Argennon application is stored in heap chunks. An application may have several heap chunks
with different sizes, and can remove or
resize its chunks or allocate new chunks. Every chunk belongs to exactly one application. Only the owner application can
modify a chunk but there is no restrictions for reading a chunk \footnote{The reason behind this type of access
control design is the fact that smart contract
code is usually immutable. That means if a smart contract does not implement a
getter mechanism for some parts of its internal data, this functionality can never
be added later, and despite the internal data is publicly available, there will be no
way for other smart contracts to use this data on-chain.}.

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself, provided it has a correct format. This
is an important feature of the AscEE heap, which allows applications to use the AscEE heap as a dictionary
data structure \footnote{also called a map.}.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily find unique identifiers for its chunks.

A heap chunk can be considered as a continuous
array of bytes. Every chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated during the session like a normal memory location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the blockchain and is proposed by the block
proposer. \texttt{sizeUpperBound} indicates the upper bound of \texttt{chunkSize} value for a chunk in a block.
A validator can use this value to safely allocate memory for a chunk.

The value of \texttt{chunkSize} at the end of the execution session will determine if a memory location at an
offset is persistent or not: Offsets lower than the chunk size are persistent, and higher offsets are not.
Non-persistent locations will be re-initialized with zero at the start of every execution session.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is defined on a chunk
and has an offset and a size and determines accessible memory locations inside the chunk. Multiple access blocks can
be defined on a single chunk, but they must be non overlapping. Access blocks are byte addressable and can have
different access types:

\begin{itemize}
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{additive}: only allows addition-like operations without overflow checking. Note that the content of
    these access blocks cannot be read.
\end{itemize}

Chunks and access blocks have different purposes. Chunks are intended for simplifying proof checking of the data
stored in the Argennon cloud \footnote{Note that
the Argennon cloud uses cryptographic accumulators for preserving data integrity.} and access blocks are
required for better parallelization of the request execution. An application should put the data it predicts is
needed for validating a block in the same chunk and the data it predicts is needed in a single execution
session in the same access block.

@@

An application can load any chunk with a valid prefix identifier even if that chunk does not exist. For a non-existent
chunk the value of \texttt{chunkSize} is always zero.

@@

The attachment of a request is a list of request identifiers of the current block that are ``attached'' to the request.
That means, for validating that request a validator first needs to ``inject'' the digest of attached requests into the
HTTP request. By doing so, the called application will have access to the digest of attachments in a secure way.

The main usage of this feature is for fee payment. A request that wants to pay the fees for a number of requests,
declares those requests as its attachments. For paying fees the payer signs the digest of requests for which he
wants to pay fees. After injecting the digest of those request by validators, that signature can be validated
correctly and securely by the called application.

@@

The AscEE uses \emph{authenticated message passing} for authorizing operations. In this method, every execution
session has a set of authenticated messages, and applications explicitly pass those messages for authorizing operations.
Authentication of messages can be efficiently done in parallel, using any type of cryptographic signature scheme in
the preprocessing phase of request execution. For example, we could use BLS aggregate signature to authenticate all
messages of a block in bulk.

Moreover, applications can use built-in functions of AscEE to generate authenticated messages in run-time. This enables
an application to authorize operations even for applications that are not directly called.

@@

Authorization by explicit messages and signatures eliminates the need for approval methods or call back patterns in
Argennon.

\note{The AscEE has no instructions for issuing cryptographic signatures.}

@@

The Argennon Smart Contract Execution Environment provides optional low level reentrancy protection by providing low
level \emph{entrance locks}. When an application acquires an entrance lock it cannot acquire that lock again and trying
to do so will result in a revert. The entrance lock of an application will be released when the application explicitly
releases its lock or when the call that had acquired that lock completes.

@@

Obviously resource monitoring is easier and more efficient for optimistic sessions. However, if
an optimistic session violates its resource caps, determining the point of failure requires precise resource
measurement. Note that for implementing optional decoupling principle, we need to determine the exact application
which has failed in a call chain. For example, assume that an optimistic session containing an application call
violates a 2 milliseconds execution time cap. If the call happens exactly after 2 milliseconds
in the caller application, a small fluctuation in the execution time measurement can
change the point of failure between the called and the main application. This can introduce nondeterministic
behaviour which could make block validation impossible. That's why monitored sessions exist. For a monitored session
this is not an issue because resource measurements and caps are defined per application call. This way, we can make
sure that resource caps are not too small and can be measured easily.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    is the amount of cpu time that is required for executing a session or an application call. The execution time is
    measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of \textbf{cpu time} needed for executing a predefined standard application which is used for benchmarking a host's
    performance.

@@

    Monitored sessions perform per application call cpu-time measurement, and every application call during a monitored
    session has a separate \texttt{maxClocks} value. This value determines the maximum amount of time that the cpu
    can be used for executing that particular application call. It should be noted that the cpu timer of the
    application call is paused when the application makes a call to another application, and is resumed when the
    control returns. An application call
    needs at least 100 clocks and if the value of its \texttt{maxClock} is lower than this value the call will be
    considered a failed call.

@@

    Optimistic sessions measure local memory usage per session and enforce a protocol-defined cap on the total amount
    of local memory a session can use. Monitored sessions measure local memory usage per
    application call and enforce a protocol-defined cap per application call. An application call which tries to use
    more local memory than the cap, will fail.
    \item \textbf{heap access list}:
    every session can only access heap locations that are declared in its access list. In addition,
    resizing heap chunks can only be done in the range of the pre-declared lower bound and upper bound.
    \item \textbf{applications list}:
    a session may only make requests to applications that are declared in its application list.
    \item \textbf{call depth}:
    during a session the number of nested application calls can not be more than a threshold. This threshold is
    determined by the Argennon protocol.
    \item \textbf{active differed calls}:
    \item \textbf{virtual signatures}:
    \item \textbf{number of entrance locks}:
\end{itemize}

The Argennon protocol allows an AscEE's implementation to have at max \%50 error in measuring execution time and at
max \%25 error in measuring local memory usage. Other resources must be measured precisely.
